---
title: "Matching Methods"
subtitle: "Covariate Matching"
bibliography: references.bib
---

# Introduction

Covariate matching methods directly match units based on the similarity of their observed characteristics, rather than reducing them to a single propensity score. These methods preserve the **multivariate structure** of covariates and can achieve better balance.

## Advantages Over Propensity Score Matching

1. **No information loss**: Matching on full covariate space preserves all information
2. **Better balance**: Can achieve exact or near-exact balance on all covariates
3. **Transparency**: Easier to inspect and understand matches
4. **Robustness**: Not sensitive to propensity score model specification

# Mahalanobis Distance Matching

## Definition

The Mahalanobis distance between units $i$ and $j$ is:

$$
D_M(X_i, X_j) = \sqrt{(X_i - X_j)^\top S^{-1} (X_i - X_j)}
$$

where $S$ is the sample covariance matrix of covariates.

## Why Not Euclidean Distance?

Euclidean distance treats all variables equally:

$$
D_E(X_i, X_j) = \sqrt{(X_i - X_j)^\top (X_i - X_j)}
$$

**Problems:**

- Variables with larger scales dominate the distance
- Ignores correlation structure among covariates

Mahalanobis distance **standardizes** by $S^{-1}$, making it:

- **Scale-invariant**: Variables measured in different units are comparable
- **Correlation-adjusted**: Accounts for dependencies among covariates

## Properties

1. **Invariant under affine transformations**: Distance unchanged if covariates are linearly transformed
2. **Multivariate**: Operates in full covariate space
3. **Non-parametric**: No model for treatment assignment required

## Limitations

- **Curse of dimensionality**: Performance degrades with many covariates
- **Sensitive to outliers**: Extreme values can distort $S$
- **Assumes multivariate normality** (implicitly)

## Implementation in R

```{r}
#| eval: false

library(MatchIt)
library(Matching)

# Mahalanobis distance matching
m.out <- matchit(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = mydata,
  method = "nearest",
  distance = "mahalanobis"
)

# With caliper on propensity score
m.out2 <- matchit(
  treat ~ age + educ + race + married + nodegree + re74 + re75,
  data = mydata,
  method = "nearest",
  distance = "mahalanobis",
  mahvars = ~ age + educ + re74 + re75,  # Variables for Mahal distance
  caliper = 0.25,                         # Caliper on propensity score
  std.caliper = TRUE
)

# Check balance
summary(m.out, un = FALSE)
```

## Implementation in Python

```{python}
#| eval: false

import numpy as np
from scipy.spatial.distance import mahalanobis
from sklearn.covariance import EmpiricalCovariance

# Compute covariance matrix
X = data[['age', 'educ', 'married', 're74', 're75']].values
cov = EmpiricalCovariance().fit(X)
S_inv = np.linalg.inv(cov.covariance_)

# Calculate distance between two units
dist = mahalanobis(X[i], X[j], S_inv)

# Using causalinference package
from causalinference import CausalModel

Y = data['outcome'].values
D = data['treat'].values
X = data[['age', 'educ', 'married', 're74', 're75']].values

causal = CausalModel(Y, D, X)
causal.est_via_matching(matches=1, bias_adj=True)
print(causal.estimates)
```

## Implementation in Stata

```stata
* Install mahapick if needed
* ssc install mahapick

mahapick treat age educ black hispanic married nodegree re74 re75, ///
         idvar(id) nummatches(1) scorevar(mahal_dist) matchedvar(matched)

* Keep matched observations
keep if matched == 1

* Estimate treatment effect
reg outcome treat age educ black hispanic married nodegree re74 re75
```

# Coarsened Exact Matching (CEM)

## Concept

CEM works by:

1. **Coarsening** continuous variables into discrete bins
2. **Exact matching** on coarsened variables
3. **Pruning** strata with no treatment or control units
4. **Analysis** on original (uncoarsened) data

## Algorithm

**Input:** Treatment $T$, covariates $X$, outcome $Y$

1. Define coarsening scheme $C(X)$ (e.g., age → age groups)
2. Sort units into strata based on $C(X)$
3. Discard strata containing only treated or only control units
4. Compute stratum weights to create balance
5. Estimate treatment effect using weighted data

## Monotonic Imbalance Bounding

CEM guarantees that **finer coarsening cannot increase imbalance**.

Let $L(C(X))$ measure imbalance under coarsening $C$. If $C_1 \preceq C_2$ (finer), then:

$$
L(C_1(X)) \leq L(C_2(X))
$$

This provides **ex ante control** over balance.

## Implementation in R

```{r}
#| eval: false

library(cem)

# Automatic coarsening
mat <- cem(
  treatment = "treat",
  data = mydata,
  drop = "outcome",
  keep.all = TRUE
)

# Manual coarsening with custom bins
educut <- c(0, 6.5, 8.5, 12.5, 17)
mat2 <- cem(
  treatment = "treat",
  data = mydata,
  drop = "outcome",
  cutpoints = list(education = educut)
)

# Check imbalance
imbalance(group = mydata$treat, data = mydata[covariates])

# Estimate treatment effect with weights
fit <- lm(outcome ~ treat + age + educ, data = mydata, weights = mat$w)
summary(fit)
```

## Implementation in Python

```{python}
#| eval: false

# CEM not natively available in Python
# Use rpy2 to call R's cem package

import rpy2.robjects as ro
from rpy2.robjects import pandas2ri
pandas2ri.activate()

ro.r('library(cem)')

# Convert pandas df to R df
r_data = pandas2ri.py2rpy(data)

# Run CEM in R
result = ro.r.cem('treat', data=r_data, drop='outcome')
weights = ro.r('result$w')
```

## Implementation in Stata

```stata
* Install cem
* ssc install cem

cem age (#3) educ (#4) black hispanic married, treatment(treat)

* Check balance
imbalance age educ black hispanic married, by(treat)

* Regression with CEM weights
reg outcome treat age educ black hispanic married [iw=cem_weights]
```

# Genetic Matching

## Concept

Genetic matching **optimizes the weight matrix** $W$ in a generalized Mahalanobis distance to maximize covariate balance:

$$
D_{\text{gen}}(X_i, X_j) = \sqrt{(X_i - X_j)^\top W (X_i - X_j)}
$$

where $W$ is **learned** using a genetic algorithm.

## Genetic Algorithm

1. **Initialize** population of weight matrices
2. **Evaluate** balance for each weight matrix
3. **Select** best-performing matrices ("fittest")
4. **Crossover** and **mutate** to create new generation
5. **Repeat** until convergence

## Balance Criteria

Minimize:

$$
B = \sum_{k=1}^p w_k b_k^2
$$

where $b_k$ is a balance metric (e.g., standardized mean difference, KS statistic) for covariate $k$.

## Implementation in R

```{r}
#| eval: false

library(Matching)

# Define covariates
X <- cbind(age, educ, black, hisp, married, nodegr, u74, u75, re75, re74)

# Genetic matching to find optimal weights
genout <- GenMatch(
  Tr = treat,
  X = X,
  estimand = "ATT",
  M = 1,
  pop.size = 100,
  max.generations = 25
)

# Match using optimized weights
mout <- Match(
  Y = outcome,
  Tr = treat,
  X = X,
  estimand = "ATT",
  Weight.matrix = genout
)

# Results
summary(mout)

# Assess balance
mb <- MatchBalance(
  treat ~ age + educ + black + hisp + married + nodegr + u74 + u75 + re75 + re74,
  match.out = mout,
  nboots = 500
)
```

# Entropy Balancing

## Concept

Entropy balancing **reweights** control units to exactly match treatment group moments while minimizing information loss (maximum entropy).

## Optimization Problem

Find weights $\{w_i\}_{i:T_i=0}$ that:

1. **Balance constraints:**
$$
\sum_{i:T_i=0} w_i X_i = \bar{X}_T
$$

2. **Minimize KL divergence:**
$$
\min \sum_{i:T_i=0} w_i \log(w_i / w_i^{(0)})
$$

where $w_i^{(0)} = 1/n_0$ are uniform base weights.

## Advantages

- **Exact balance** on specified moments (means, variances, etc.)
- **Unique solution** via convex optimization
- **No iteration** required

## Implementation in R

```{r}
#| eval: false

library(ebal)

# Entropy balancing
eb.out <- ebalance(
  Treatment = treat,
  X = cbind(age, educ, black, married, re74, re75)
)

# Construct weights
weights <- c(eb.out$w, rep(1, sum(treat)))

# Estimate ATE
fit <- lm(outcome ~ treat, weights = weights)
summary(fit)
```

# Comparison of Methods

| **Method** | **Distance** | **Pros** | **Cons** |
|------------|-------------|----------|----------|
| Mahalanobis | Multivariate | Scale-invariant, preserves structure | High-dimensional issues |
| CEM | Exact (discretized) | Monotonic balance, interpretable | Information loss from coarsening |
| Genetic | Optimized | Data-adaptive, superior balance | Computationally intensive |
| Entropy | Reweighting | Exact balance, efficient | Requires all units, potential extreme weights |

::: {#refs}
:::
