---
title: "Coding Practice"
bibliography: references.bib
---

Now that you have a sense of the purpose and methodology behind matching methods, we can attempt a simple application of coding this in practice. You are welcome to use your coding language of preference (`Stata`, `R`, or `Python`). But I will presume that you have both established your working environment to execute the code and familiarity in the language to adequately troubleshoot your technical issues. 


# Naive Estimation 

First, estimate the treatment effect **without any matching/adjustment**. This gives a baseline estimate that may be biased by selection on observables.

The naive estimator is:

$$
\hat\tau_{naive} = E[Y\mid T=1] - E[Y\mid T=0]
$$

If treated and control groups differ systematically in pre-treatment covariates, this difference combines treatment effect plus selection bias.

::: {.panel-tabset}

## R

```{r}
#| eval: false

library(causaldata)
library(dplyr)

br <- black_politicians

# Naive difference in means
naive_diff <- with(br, mean(responded[leg_black == 1]) - mean(responded[leg_black == 0]))
naive_diff

# Naive OLS (unadjusted)
naive_lm <- lm(responded ~ leg_black, data = br)
summary(naive_lm)
```

## Python

```{python}
#| eval: false

from causaldata import black_politicians
import statsmodels.formula.api as smf

br = black_politicians.load_pandas().data

naive_diff = br.loc[br.leg_black == 1, "responded"].mean() - \
                         br.loc[br.leg_black == 0, "responded"].mean()
print("Naive diff in means:", naive_diff)

naive_model = smf.ols("responded ~ leg_black", data=br).fit()
print(naive_model.summary())
```

## Stata

```{stata}
#| eval: false

* Naive difference in means
mean responded, over(leg_black)

* Naive unadjusted regression
reg responded leg_black
```

:::

# Matching Estimation 

Now compare with the matching-based estimate from the code above. The key teaching point is:

- If $\hat\tau_{naive}$ and $\hat\tau_{match}$ differ meaningfully, imbalance likely biased the naive estimate.

Define an estimated bias proxy:

$$
\widehat{Bias}_{selection} \approx \hat\tau_{naive} - \hat\tau_{match}
$$

This is not the true bias (which is unobserved), but it is a useful diagnostic for teaching.

::: {.panel-tabset}

## R

```{r}
#| eval: false

# From previous sections:
# naive_diff already computed
# M <- Match(...)

match_est <- M$est
bias_proxy <- naive_diff - match_est

cat("Naive estimate:", naive_diff, "\n")
cat("Matching estimate:", match_est, "\n")
cat("Estimated selection bias proxy (naive - match):", bias_proxy, "\n")
```

## Stata

```{stata}
#| eval: false

* Save naive coefficient
reg responded leg_black
scalar b_naive = _b[leg_black]

* Matching estimator (ATT reported by teffects)
teffects nnmatch (responded medianhhincom blackpercent leg_democrat) ///
        (leg_black), nneighbor(1) metric(mahalanobis)

matrix b = e(b)
scalar b_match = b[1,1]

display "Naive estimate: " b_naive
display "Matching estimate: " b_match
display "Estimated selection bias proxy (naive - match): " (b_naive - b_match)
```

## Python

```{python}
#| eval: false

# naive_diff from previous section
# M.estimates from causalinference output

print(M.estimates)
print("Compare naive_diff to ATT/ATE above to discuss selection bias.")
```

:::

# Sensitivity Analysis 

Even after matching, hidden confounding may remain. Use sensitivity analysis to ask:
"How strong would an unobserved confounder need to be to overturn the result?"

For matched pairs, Rosenbaum bounds are standard.

::: {.panel-tabset}

## R

```{r}
#| eval: false

library(MatchIt)
library(rbounds)

# Reconstruct matched dataset via MatchIt for sensitivity workflow
m.out <- matchit(
    leg_black ~ medianhhincom + blackpercent + leg_democrat,
    data = br,
    method = "nearest",
    ratio = 1
)

md <- match.data(m.out)

# Build paired outcomes by subclass (1 treated + 1 control each subclass)
pair_summ <- md %>%
    group_by(subclass) %>%
    summarize(
        y_t = responded[leg_black == 1][1],
        y_c = responded[leg_black == 0][1],
        .groups = "drop"
    )

# Rosenbaum sensitivity for Wilcoxon signed-rank p-value
psens(pair_summ$y_t, pair_summ$y_c, Gamma = 2, GammaInc = 0.1)

# Hodges-Lehmann sensitivity
hlsens(pair_summ$y_t, pair_summ$y_c, Gamma = 2, GammaInc = 0.1)
```

Interpretation guide:

- If significance disappears around $\Gamma \approx 1.1$, results are fragile.
- If significance persists until $\Gamma > 2$, results are relatively robust.


## Python

```{python}
#| eval: false

# Python has limited native Rosenbaum-bounds tooling.
# Practical option: run sensitivity in R (rbounds) after exporting matched pairs,
# or use rpy2 from Python to call psens()/hlsens().
```

## Stata

```{stata}
#| eval: false

* After nearest-neighbor matching, run rbounds (user-written)
* ssc install rbounds

* Example sensitivity check across gamma values
rbounds responded, gamma(1(0.1)2)
```


:::

Suggested student exercise:

1. Compute and report naive and matched estimates.
2. Quantify the bias proxy (naive - matched).
3. Run Rosenbaum bounds and report the critical Gamma where inference changes.
4. Conclude whether results are robust to moderate hidden bias.


# References

::: {#refs}
:::
