---
title: "Coding Practice"
bibliography: references.bib
---

Now that you have a sense of the purpose and methodology behind matching methods, we can attempt a simple application of coding this in practice. You are welcome to use your coding language of preference (`Stata`, `R`, or `Python`). But I will presume that you have both established your working environment to execute the code and familiarity in the language to adequately troubleshoot your technical issues. 


For this exercise, we will be using data from XXX. You can download the sample data set here [](). 


::: {.panel-tabset}

## R

```{r}
#| eval: false
#| cache: true

library(Matching); library(tidyverse)
br <- causaldata::black_politicians

# Outcome
Y <- br %>%
    pull(responded)
# Treatment
D <- br %>%
    pull(leg_black)
# Matching variables
# Note select() is also in the Matching package, so we specify dplyr
X <- br %>%
    dplyr::select(medianhhincom, blackpercent, leg_democrat) %>%
    as.matrix()

# Weight = 2, oddly, denotes Mahalanobis distance
M <- Match(Y, D, X, Weight = 2, caliper = 1)

# See treatment effect estimate
summary(M)

# Get matched data for use elsewhere. Note that this approach will 
# duplicate each observation for each time it was matched
matched_treated <- tibble(id = M$index.treated,
                          weight = M$weights)
matched_control <- tibble(id = M$index.control,
                          weight = M$weights)
matched_sets <- bind_rows(matched_treated,
                          matched_control) 
# Simplify to one row per observation
matched_sets <- matched_sets %>%
                    group_by(id) %>%
                    summarize(weight = sum(weight))
# And bring back to data
matched_br <- br %>%
    mutate(id = row_number()) %>%
    left_join(matched_sets, by = 'id')

# To be used like this! The standard errors here are wrong
lm(responded~leg_black, data = matched_br, weights = weight)
```

## Stata

```{stata}
#| eval: false
#| cache: true

causaldata black_politicians.dta, use clear download

* Create an id variable based on row number,
* which will be used to locate match IDs
g id = _n

* teffects nnmatch does nearest-neighbor matching using first the outcome 
* variable and the matching variables then the treatment variable.
* The generate option creates variables with the match identities
teffects nnmatch (responded medianhhincom blackpercent leg_democrat) ///
    (leg_black), nneighbor(1) metric(mahalanobis) gen(match)

* Note this will produce a treatment effect by itself
* Doing anything else with the match in Stata is pretty tricky
* (or using a caliper). If you want either of those you may want to 
* look into the mahapick or psmatch2 packages
* Although these won't give the most accurate standard errors
```


## Python

```{python}
#| eval: false
#| cache: true

import pandas as pd
import numpy as np
# The more-popular matching tools in sklearn
# are more geared towards machine learning than statistical inference
from causalinference.causal import CausalModel
from causaldata import black_politicians
br = black_politicians.load_pandas().data

# Get our outcome, treatment, and matching variables
# We need these as numpy arrays
Y = br['responded'].to_numpy()
D = br['leg_black'].to_numpy()
X = br[['medianhhincom', 'blackpercent', 'leg_democrat']].to_numpy()

# Set up our model
M = CausalModel(Y, D, X)

# Fit, using Mahalanobis distance
M.est_via_matching(weights = 'maha', matches = 1)

print(M.estimates)
# Note it automatically calcultes average treatments on
# average, on treated, and on untreated/control (ATC)
```

:::





# Naive Estimation 

First, estimate the treatment effect **without any matching/adjustment**. This gives a baseline estimate that may be biased by selection on observables.

The naive estimator is:

$$
\hat\tau_{naive} = E[Y\mid T=1] - E[Y\mid T=0]
$$

If treated and control groups differ systematically in pre-treatment covariates, this difference combines treatment effect plus selection bias.

::: {.panel-tabset}

## R

```{r}
#| eval: false

library(causaldata)
library(dplyr)

br <- black_politicians

# Naive difference in means
naive_diff <- with(br, mean(responded[leg_black == 1]) - mean(responded[leg_black == 0]))
naive_diff

# Naive OLS (unadjusted)
naive_lm <- lm(responded ~ leg_black, data = br)
summary(naive_lm)
```

## Stata

```{stata}
#| eval: false

* Naive difference in means
mean responded, over(leg_black)

* Naive unadjusted regression
reg responded leg_black
```

## Python

```{python}
#| eval: false

from causaldata import black_politicians
import statsmodels.formula.api as smf

br = black_politicians.load_pandas().data

naive_diff = br.loc[br.leg_black == 1, "responded"].mean() - \
                         br.loc[br.leg_black == 0, "responded"].mean()
print("Naive diff in means:", naive_diff)

naive_model = smf.ols("responded ~ leg_black", data=br).fit()
print(naive_model.summary())
```

:::

# Matching Estimation 

Now compare with the matching-based estimate from the code above. The key teaching point is:

- If $\hat\tau_{naive}$ and $\hat\tau_{match}$ differ meaningfully, imbalance likely biased the naive estimate.

Define an estimated bias proxy:

$$
\widehat{Bias}_{selection} \approx \hat\tau_{naive} - \hat\tau_{match}
$$

This is not the true bias (which is unobserved), but it is a useful diagnostic for teaching.

::: {.panel-tabset}

## R

```{r}
#| eval: false

# From previous sections:
# naive_diff already computed
# M <- Match(...)

match_est <- M$est
bias_proxy <- naive_diff - match_est

cat("Naive estimate:", naive_diff, "\n")
cat("Matching estimate:", match_est, "\n")
cat("Estimated selection bias proxy (naive - match):", bias_proxy, "\n")
```

## Stata

```{stata}
#| eval: false

* Save naive coefficient
reg responded leg_black
scalar b_naive = _b[leg_black]

* Matching estimator (ATT reported by teffects)
teffects nnmatch (responded medianhhincom blackpercent leg_democrat) ///
        (leg_black), nneighbor(1) metric(mahalanobis)

matrix b = e(b)
scalar b_match = b[1,1]

display "Naive estimate: " b_naive
display "Matching estimate: " b_match
display "Estimated selection bias proxy (naive - match): " (b_naive - b_match)
```

## Python

```{python}
#| eval: false

# naive_diff from previous section
# M.estimates from causalinference output

print(M.estimates)
print("Compare naive_diff to ATT/ATE above to discuss selection bias.")
```

:::

# Sensitivity Analysis 

Even after matching, hidden confounding may remain. Use sensitivity analysis to ask:
"How strong would an unobserved confounder need to be to overturn the result?"

For matched pairs, Rosenbaum bounds are standard.

::: {.panel-tabset}

## R

```{r}
#| eval: false

library(MatchIt)
library(rbounds)

# Reconstruct matched dataset via MatchIt for sensitivity workflow
m.out <- matchit(
    leg_black ~ medianhhincom + blackpercent + leg_democrat,
    data = br,
    method = "nearest",
    ratio = 1
)

md <- match.data(m.out)

# Build paired outcomes by subclass (1 treated + 1 control each subclass)
pair_summ <- md %>%
    group_by(subclass) %>%
    summarize(
        y_t = responded[leg_black == 1][1],
        y_c = responded[leg_black == 0][1],
        .groups = "drop"
    )

# Rosenbaum sensitivity for Wilcoxon signed-rank p-value
psens(pair_summ$y_t, pair_summ$y_c, Gamma = 2, GammaInc = 0.1)

# Hodges-Lehmann sensitivity
hlsens(pair_summ$y_t, pair_summ$y_c, Gamma = 2, GammaInc = 0.1)
```

Interpretation guide:

- If significance disappears around $\Gamma \approx 1.1$, results are fragile.
- If significance persists until $\Gamma > 2$, results are relatively robust.

## Stata

```{stata}
#| eval: false

* After nearest-neighbor matching, run rbounds (user-written)
* ssc install rbounds

* Example sensitivity check across gamma values
rbounds responded, gamma(1(0.1)2)
```

## Python

```{python}
#| eval: false

# Python has limited native Rosenbaum-bounds tooling.
# Practical option: run sensitivity in R (rbounds) after exporting matched pairs,
# or use rpy2 from Python to call psens()/hlsens().
```

:::

Suggested student exercise:

1. Compute and report naive and matched estimates.
2. Quantify the bias proxy (naive - matched).
3. Run Rosenbaum bounds and report the critical Gamma where inference changes.
4. Conclude whether results are robust to moderate hidden bias.

::: {#refs}
:::
