---
title: "Sensitivity Analysis"
bibliography: references.bib
---

Matching estimates are only valid if the **ignorability assumption** holds—that all confounders are observed. Since this assumption is untestable, sensitivity analyses assess how robust the results are to potential violations.

# Assessing Common Support

## Definition

**Common support** (or overlap) requires that for every covariate value, there is a positive probability of receiving both treatment and control:

$$
0 < P(T = 1 | X = x) < 1 \quad \forall x
$$

Without overlap, comparisons require **extrapolation** beyond the empirical data, making causal inference unreliable.

## Diagnosing Lack of Overlap

### 1. Propensity Score Histograms

```{r}
#| eval: false

library(ggplot2)

# Visualize propensity score distributions
ggplot(data, aes(x = pscore, fill = factor(treat))) +
  geom_histogram(alpha = 0.5, position = "identity", bins = 30) +
  labs(x = "Propensity Score", y = "Count", fill = "Treatment") +
  theme_minimal()
```

**Look for:**

- Non-overlapping regions at extreme values
- Treatment group with no control counterparts (or vice versa)

### 2. Convex Hull Method

The **convex hull** identifies the multidimensional space where interpolation (not extrapolation) is possible [@king2006dangers].

- Discard units outside the convex hull of the opposite group
- More restrictive than propensity score trimming

### 3. Discarding Units Outside Common Support

**By propensity score:**

```{r}
#| eval: false

# Define range
ps_min <- max(min(data$pscore[data$treat == 1]), 
            min(data$pscore[data$treat == 0]))
ps_max <- min(max(data$pscore[data$treat == 1]), 
            max(data$pscore[data$treat == 0]))

# Keep only units in common support
data_trimmed <- data %>%
  filter(pscore >= ps_min & pscore <= ps_max)
```

## Implications for Estimands

- **ATT**: It may be acceptable to discard control units outside the treated range
- **ATE**: Discarding **any** units changes the target population and reduces external validity

# Balance Diagnostics

After matching, verify that covariates are balanced between treated and control groups.

## Standardized Mean Differences

The **standardized mean difference** (SMD) for covariate $X$ is:

$$
\text{SMD} = \frac{\bar{X}_T - \bar{X}_C}{\sqrt{\frac{s_T^2 + s_C^2}{2}}}
$$

where $\bar{X}_T, \bar{X}_C$ are means and $s_T^2, s_C^2$ are variances in treated and control groups.

**Benchmark:** SMD $< 0.1$ indicates good balance [@rubin2001using]

## Variance Ratios

Compare variances of continuous covariates:

$$
\text{VR} = \frac{s_T^2}{s_C^2}
$$

**Benchmark:** $0.5 < \text{VR} < 2.0$ [@rubin2001using]

## Propensity Score Balance

**Standardized difference** of propensity score means: Should be $< 0.25$  
**Variance ratio** of propensity scores: Should be between 0.5 and 2.0

## Graphical Diagnostics

### Love Plots

Display standardized mean differences before and after matching:

```{r}
#| eval: false

library(cobalt)

bal.plot(m.out, var.name = "distance", which = "both")

love.plot(m.out, threshold = 0.1, 
        abs = TRUE, 
        var.order = "unadjusted")
```

### QQ Plots

Compare empirical distributions of covariates:

```{r}
#| eval: false

plot(m.out, type = "qq", which.xs = c("age", "educ", "re74"))
```

### Empirical CDFs

```{r}
#| eval: false

plot(m.out, type = "ecdf")
```

## Avoid Using P-values

**Do not** use t-tests or other hypothesis tests to assess balance:

- P-values conflate **effect size** with **sample size**
- Large samples will show "significant" differences even for trivial imbalances
- Balance is a **design** property, not a hypothesis test

# Rosenbaum Sensitivity Analysis

## Motivation

Even after matching, **hidden bias** from unobserved confounders $U$ may remain. Rosenbaum bounds quantify how strong this bias must be to overturn the findings.

## Setup

Let $\pi_i = P(T_i = 1)$ be the true probability of treatment for unit $i$. Define the **odds** of treatment:

$$
\text{Odds}_i = \frac{\pi_i}{1 - \pi_i}
$$

For two matched units $i$ and $j$, the **odds ratio** is bounded by $\Gamma$:

$$
\frac{1}{\Gamma} \leq \frac{\text{Odds}_i}{\text{Odds}_j} \leq \Gamma
$$

## Interpretation

- $\Gamma = 1$: No hidden bias (treatment is "as good as randomly assigned" conditional on $X$)
- $\Gamma = 2$: One unit could be **twice as likely** to receive treatment due to an unobserved confounder
- $\Gamma = 3$: Three times as likely

## Procedure

1. Choose a range of $\Gamma$ values (e.g., 1.0, 1.1, 1.2, ..., 3.0)
2. For each $\Gamma$, compute **bounds** on the p-value or treatment effect estimate
3. Report the **critical $\Gamma$** at which the effect becomes insignificant

::: {.panel-tabset}

## R

```{r}
#| eval: false

library(rbounds)
library(MatchIt)

# Perform matching
m.out <- matchit(treat ~ age + educ, data = mydata, method = "nearest")
matched_data <- match.data(m.out)

# Split into groups
treat_y <- matched_data$outcome[matched_data$treat == 1]
control_y <- matched_data$outcome[matched_data$treat == 0]

# P-value sensitivity
psens(treat_y, control_y, Gamma = 3, GammaInc = 0.1)

# Hodges-Lehmann point estimate sensitivity
hlsens(treat_y, control_y, Gamma = 3, GammaInc = 0.1)
```

### Output

- Tables showing p-values and confidence intervals at different $\Gamma$ levels
- Identify $\Gamma$ where effect loses significance

### Interpretation

If the critical $\Gamma$ is:

- **< 1.5**: Results are **very sensitive** to hidden bias
- **1.5 - 2.5**: Moderately robust
- **> 2.5**: Highly robust—unobserved confounder would need to be very strong

## Stata

```{stata}
#| eval: false

* After matching, save matched sample
ssc install rbounds

* Sensitivity analysis
rbounds outcome, gamma(1(0.1)3)
```

## Python

```{python}
#| eval: false

# No direct Rosenbaum bounds in Python
# Use R via rpy2 or implement manually using Wilcoxon signed-rank test
import rpy2.robjects as ro
ro.r('library(rbounds)')
result = ro.r('psens(treat_y, control_y, Gamma=2)')
```

:::

# Relative Correlation Restrictions

An alternative sensitivity approach bounds the treatment effect under assumptions about **selection on unobservables** relative to **selection on observables** [@altonji2005selection].

## Framework

Assume:

$$
\text{Cov}(T, \epsilon) = \lambda \cdot \text{Cov}(T, X^\top \gamma)
$$

where:

- $\epsilon$ contains unobserved confounders
- $\lambda$ is the **relative selection ratio**

## Interpretation

- $\lambda = 0$: No selection on unobservables (standard OLS)
- $\lambda = 1$: Selection on unobservables is **equal** to selection on observables
- $\lambda > 1$: Unobservables are **more important** than observables

**Critical value $\lambda^*$**: The value at which the treatment effect reaches zero

## Implementation

```{r}
#| eval: false

library(rcrbounds)

# Estimate bounds
rcr_result <- rcr(
  formula = outcome ~ treat | age + educ + married,
  data = mydata,
  rc_range = c(0, 10)
)

summary(rcr_result)
plot(rcr_result)
```

# Coefficient Stability (Oster's delta)

@oster2019unobservable proposes testing robustness by examining how coefficients change as controls are added.

## Key Metrics

1. **Bias-adjusted coefficient** beta_star(delta, Rmax^2)
2. **Breakdown point** delta_star: Value of delta at which beta_star = 0

**Interpretation:**

- delta_star > 1: Unobservables would need to be **more selective** than observables to eliminate the effect
- delta_star < 1: Results fragile to hidden confounding

## Implementation

```{r}
#| eval: false

library(robomit)

# Compute delta*
delta_result <- o_delta(
  y = "outcome",
  x = "treat",
  con = "age + educ + married + re74 + re75",
  beta = 0,
  R2max = 1.3 * summary(lm_model)$r.squared,
  type = "lm",
  data = mydata
)

# Visualize
o_delta_rsq_viz(
  y = "outcome", x = "treat",
  con = "age + educ + married + re74 + re75",
  beta = 0,
  data = mydata
)
```

# Falsification Tests

## Placebo Outcomes

Test whether the treatment "affects" outcomes it **should not** affect (e.g., pre-treatment variables, irrelevant outcomes).

```{r}
#| eval: false

# Match as usual
m.out <- matchit(treat ~ covariates, data = mydata)

# Test on placebo outcome
placebo_model <- lm(pre_treatment_outcome ~ treat, 
                    data = match.data(m.out),
                    weights = weights)

# Should find no effect
summary(placebo_model)
```

## Permutation Tests

Randomly reassign treatment status and re-estimate the effect. The true effect should be an outlier in the permutation distribution.

# Untestable Limitations

Some assumptions **cannot be tested** empirically:

1. **Unconfoundedness**: We can never prove all confounders are observed
2. **SUTVA**: Interference effects are hard to detect directly
3. **Functional form**: Matching is somewhat robust, but outcome models still require assumptions

**Strategies:**

- **Domain knowledge**: Use theory to justify covariate selection
- **Sensitivity analyses**: Quantify how robust conclusions are
- **Robustness checks**: Try multiple matching methods and specifications
- **Transparency**: Report all decisions and diagnostics


# References

::: {#refs}
:::
